Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Cleaned up technique section, moved to experiment section

\subsection*{The Experiment}
We construct an experiment to test the efficacy of a predictive model. 
For either similarity metric, there is not enough data to actually calculate
$TrueSim$ for any pair. Instead, we can get close by calculating the
corresponding $Sim_n$ metric when a large number of users have rated both 
items. Formally, we define $GoldSim$ = $Sim_N$ where $N$ is a number large 
enough that $GoldSim \approx TrueSim$.

With a measurable $GoldSim$ as our predictive goal, we can implement a simple
predictivity experiment. First, we find pairs of items that have at least 
$N$ common users and calculate their $GoldSim$. Next, for each of these pairs
of items, we withhold user ratings, leaving $n$ common reviewers. This 
simulates a pair of items at times when there is a sparse amount of rating data. 
We repeat theis for $n=1,...,N$ to get varying degrees of sparsity. With the
sparse ratings alone, the model in question generates a prediction of the 
$GoldSim$, which is compared to the actual $GoldSim$. This is repeated over 
a large number of test pairs and the average squared error of the prediction 
is calculated for each $n$.

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu_s, \sigma_{s}^2),
\end{align}
where $\mu_s$ and $\sigma_{s}^2$ are the average similarity score and the variance
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{err, n}^2)
\end{align}
Note that $\sigma_{err, n}^2$ represents how noisy $Sim_n$ is as an estimate of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ is illustrated in Fig. 1.

@@

\subsection*{Technique}
To evaluate this method of improving similarity estimates, we would like to know
{\em a priori} the true similarity for a set of item pairs. This way, we could
compute the error in estimates made based on only partial rating data. Of
course, the metric $TrueSim$ is a theoretical construct and not something we
have access to in the real world. Fortunately, we can achieve good
approximations of $TrueSim$ given sufficient data. We therefore define $GoldSim$
= $Sim_N$ where $N$ is a number large enough that $GoldSim \approx TrueSim$.
Using this approach to approximate $TrueSim$, we can look at the distribution of
$GoldSim$ scores (Fig. 2) and see that the Normal distribution of $TrueSim$ is a
reasonable model.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Histograms.png}
	\caption{Histograms of {\em TrueSimR} (top) and {\em TrueSimP} (bottom) for
training datasets containing at least 40 common reviewers (left) and 80 common
reviewers (right).}
    \label{fig:Histograms}
\end{figure}

With a measurable $GoldSim$ as our predictive goal, we can apply supervised 
learning techniques. First, we find pairs of items that have at least $N$ common
users and calculate their $GoldSim$. Next we withold some of the users to 
produce estimated $Sim_n$ for $n=1,...,N$. By doing this over many pairs of 
items, we produce pairs of ($Sim_n$, $TrueSim$) across different values of $n$ 
that can be used for training and testing. Now the parameters $\mu_s$, 
$\sigma_{s}^2$, and $\sigma_{err,n}^2$ can be approximated using the method of 
moments over the training data. The values of $\sigma_{err,n}^2$ are fit to the 
function $\frac{\alpha}{\sqrt{n}}$ and the $\alpha$ that minimizes the sum of 
squared error is computed.

>>>>

\subsection*{The Experiment}
We construct an experiment to test the efficacy of a predictive model. 
We would like to know {\em a priori} the true similarity for a pair of items.
For either true similarity metric, $TrueSim$ is a theoretical construct and not 
something we have access to in the real world. Fortunately, we can achieve good
approximations of $TrueSim$ given sufficient data. We therefore define $GoldSim$
= $Sim_N$ where $N$ is a number large enough that $GoldSim \approx TrueSim$.

With a measurable $GoldSim$ as our goal, we can implement a simple
predictivity experiment. First, we find pairs of items that have at least 
$N$ common users and calculate their $GoldSim$. Next, for each of these pairs
of items, we withhold user ratings, leaving $n$ common reviewers. This 
simulates a pair of items at a time when there is sparser rating data. 
We repeat this for $n=1,...,N$ to get varying degrees of sparsity. With the
sparse ratings alone, the model in question generates a prediction of the 
$GoldSim$, which is compared to the actual $GoldSim$. This is repeated over 
a large number of test pairs and the average squared error of the prediction 
is calculated for each $n$.

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu_s, \sigma_{s}^2),
\end{align}
where $\mu_s$ and $\sigma_{s}^2$ are the average similarity score and the variance
of similarity scores of all pairs of items, respectively. Figure 1 shows the
distributions of $GoldSim$, which motivates a Gaussina model for $TrueSim$.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Histograms.png}
	\caption{Histograms of {\em GoldPearSim} (top) and {\em GoldCosSim}
(bottom) for training datasets containing at least 40 common reviewers (left)
and 80 common reviewers (right).}
    \label{fig:Histograms}
\end{figure}

Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{err, n}^2)
\end{align}
Note that $\sigma_{err, n}^2$ represents how noisy $Sim_n$ is as an estimate of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ is illustrated in Figure 2.

@@

\subsection*{Technique and Results}
With a measurable $GoldSim$ as our predictive goal, we can apply supervised 
learning techniques. First, we find pairs of items that have at least $N$ common
users and calculate their $GoldSim$. Next we withold some of the users to 
produce estimated $Sim_n$ for $n=1,...,N$. By doing this over many pairs of 
items, we produce pairs of ($Sim_n$, $TrueSim$) across different values of $n$ 
that can be used for training. Now the parameters $\mu_s$, $\sigma_{s}^2$, and 
$\sigma_{err,n}^2$ can be approximated using the method of moments over the 
training data. The values of $\sigma_{err,n}^2$ are fit to the function 
$\frac{\alpha}{\sqrt{n}}$ and the $\alpha$ that minimizes the sum of squared 
error is computed.
